---
title: "STAT3069 Final Project - R"
authors: Brian Beardsall
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
---

```{r setup}
library(tidyverse)
library(rdist)
library(viridis)
library(urbnmapr)
library(dagitty)
library(ggdag)
```

# DAG
```{r}
dag <- dagitty("dag {
             MI_c -> g_s -> G_s
             MI_c -> L_s -> G_s
             MI_c -> P_c -> SW_c

              G_s -> CP_c
              
              G_s -> SW_c
              CP_c -> SW_c
              MI_c -> SW_c
              V_c -> SW_c
               }")

plot <- ggdag(dag)+
  theme_dag()

plot
```

```{r}
ggsave('dag.png')
```


# Data Processing
Load libraries.


## Presidential data
From https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/VOQCHQ
```{r}
presRaw <- read_csv('data/countypres_2000-2016.csv')
head(presRaw)
```

## Volatility

Volatility calculated as the sum of the pairwise euclidean distance matrix, for elections from 2000-2016. Coordinates for each year were (proportionDemocrat, proportionRepublican).
```{r}
volatility <- presRaw %>%
  # compute proportion of total votes for each candidate
  mutate(proportionVotes = candidatevotes / totalvotes) %>%
  # pivot on each county and year
  pivot_wider(
    id_cols = c('FIPS', 'year'),
    names_from = party, 
    values_from = proportionVotes
  ) %>%
  # makes nested dataframe, need to unnest
  unnest() %>%
  # remove years/counties missing dem or repub data
  drop_na(democrat, republican) %>%
  select(FIPS,democrat, republican) %>%
  # group and nest dataframes of proportions by FIPS
  group_by(FIPS) %>%
  nest() %>%
  # calculate volatility as the sum of pairwise distance matrix
  mutate(volatility = sum(pdist(data[[1]]))) %>%
  select(FIPS, volatility) %>%
  # convert type and zero pad FIPS to match other datasets
  mutate_at("FIPS", as.character) %>%
  mutate_at("FIPS", str_pad, width = 5, pad = "0") 

head(volatility)
```


## 2016 & 2020 Election data
From https://github.com/favstats/USElection2020-NYT-Results.
```{r}
raw2020 <- read_csv('data/2020_USElection_County_count.csv')
head(raw2020)
```

Converted Democrat margin to be positive. Swing calculated as difference in margin from 2016 (so a positive swing would be a higher Dem margin in 2020).

```{r}
pres2020 <- raw2020 %>%
  select(fips, margin2016, margin2020) %>%
  arrange(fips) %>%
  # multiply by negative to make dem positive, repub negative
  mutate(margin2016 = -margin2016,
         margin2020 = -margin2020) %>%
  # swing towards biden in 2020
  mutate(bidenSwing = margin2020 - margin2016) %>%
  rename(FIPS = fips) %>%
  # convert type and zero pad FIPS to match other datasets
  mutate_at("FIPS", as.character) %>%
  mutate_at("FIPS", str_pad, width = 5, pad = "0")

head(pres2020)
```

## Population
From https://covid19.census.gov/datasets/population-county?selectedAttribute=B01001_001E 
```{r}
popRaw <- read_csv('data/Population_-_County.csv')
head(popRaw)
```

```{r}
pop <- popRaw %>%
  rename(totalPop = B01001_001E,
         state = GEO_PARENT_NAME) %>%
  select(FIPS_CODE, state, totalPop, POP_DENSITY) %>%
  rename(FIPS = FIPS_CODE) %>%
  mutate(FIPS = as.character(FIPS))
head(pop)
```

## Race
From https://covid19.census.gov/datasets/ace8fa8bea514d07a3139e4657b3cd9c_0/data 
```{r}
raceRaw <- read_csv('data/raceCounty.csv')
head(raceRaw)
```

```{r}
race <- raceRaw %>%
  rename(whiteAlone = B02001_002E,
         county = GEO_NAME) %>%
  select(FIPS_CODE, county, whiteAlone) %>%
  rename(FIPS = FIPS_CODE) %>%
  # convert type and zero pad FIPS to match other datasets
  mutate_at("FIPS", as.character) %>%
  mutate_at("FIPS", str_pad, width = 5, pad = "0")

head(race)
```


## Cases
From https://github.com/nytimes/covid-19-data/tree/master/live, as of October 24.
```{r}
casesRaw <- read_csv('data/cases_county.csv')
casesRaw 
```
```{r}
cases <- casesRaw %>%
  rename(FIPS = fips) %>%
  select(FIPS, cases) %>%
  # convert type and zero pad FIPS to match other datasets
  mutate_at("FIPS", as.character) %>%
  mutate_at("FIPS", str_pad, width = 5, pad = "0")
head(cases)
```



## State Governance
From https://www.ncsl.org/Portals/1/Documents/Elections/Legis_Control_2020_April%201.pdf. 
```{r}
rawState <- read_csv('data/legis_gov_2020.csv')
head(rawState)
```
Each democrat level of control = +0.5, republican = -0.5, divided or NA = 0. 
```{r}
state <- rawState %>%
  select(state, legisControl, govParty, stateControl) %>%
  
  mutate(legisScore = recode(legisControl, Dem = 0.5, Rep = -0.5, Divided = 0),
         govScore = recode(govParty, Dem = 0.5, Rep = -0.5, Divided = 0)) %>%
  replace_na(list(legisScore = 0,
                  govScore = 0)) %>%
  mutate(governanceScore = legisScore + govScore)

head(state)
```

## Join Data
```{r}
joined <- pop %>%
  left_join(volatility, by = "FIPS") %>%
  left_join(state, by = "state") %>%
  left_join(cases, by = "FIPS") %>%
  left_join(race, by = "FIPS") %>%
  left_join(pres2020, by = "FIPS") %>%
  # calculate cases per capita, and proportion white.
  mutate(casesPerCap = cases / totalPop,
         propWhite = whiteAlone / totalPop)
 
head(joined)
```

```{r}
finalData <- joined %>%
  select(FIPS, county, state, volatility, governanceScore, 
         margin2016, margin2020, bidenSwing, casesPerCap, propWhite) %>%
  drop_na()

head(finalData)
```

Save as CSV.
```{r}
write_csv(finalData, 'data/processedData.csv')
```

# Map Plotting




## Read covid data by county

```{r}
covidCounty <- read_csv('data/countyCovid.csv') %>%
  rename(county_fips = FIPS) %>%
  # convert type and zero pad FIPS to match other datasets
  mutate_at("county_fips", as.character) %>%
  mutate_at("county_fips", str_pad, width = 5, pad = "0") 

covidCounty
```



## Join with built in shapefile data
```{r}
joinedCounty <- left_join(covidCounty, counties, by = "county_fips") 
```

## Plot county map
```{r}
joinedCounty %>%
  ggplot(aes(long, lat, group = group, fill = covidSlope)) +
  geom_polygon(color = NA) +
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  labs(fill = "COVID Slope") +
  scale_color_viridis()+
  scale_fill_viridis()
```

```{r}
# Save as png
ggsave('figs/countyMap.png')
```


## State map

```{r}
# read state covid slopes
covidState <- read_csv('data/stateCovidSlope.csv')
covidState
```

```{r}
# join with built in data
state_joined <- left_join(covidState, states, by = c("state" = "state_name")) 
```

```{r}
# plot map
state_joined %>%
  ggplot(aes(long, lat, group = group, fill = medianCovidSlope)) +
  geom_polygon(color = NA) +
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  labs(fill = "COVID Slope") +
  scale_color_viridis()+
  scale_fill_viridis()
```
```{r}
# save png
ggsave('figs/stateMap.png')
```

# Posterior prediction plots

```{r}
# read data
postPredCounty <- read_csv('data/postPredCounty.csv')
postPredState <- read_csv('data/postPredState.csv')
```

## State plot
```{r}
postPredState %>%
  ggplot(aes(x = SW, y = pred, color = logCP))+ 
  geom_point(alpha = 0.4)+
  # 95% UI error bars
  geom_errorbar(aes(ymin=y_195, ymax=y_u95), width=0, size = 0.5, alpha = 0.2)+ # line of slope 1 through origin
  geom_abline(slope = 1, intercept = 0, color = 'black')+
# make square
  coord_fixed(xlim = c(-4, 4), ylim = c(-4, 4))+
  theme_bw()+ 
  scale_color_viridis()+
  labs(color = "COVID Cases Per Capita (z-score)",
  y = "Predicted swing (z-score)",
  x = "Observed swing (z-score)")
```
```{r}
# save png
ggsave('figs/statePostPred.png')
```

## County plot

```{r}
postPredCounty %>%
  ggplot(aes(x = SW, y = pred, color = logCP))+ 
  geom_point(alpha = 0.4)+
  # 95% UI error bars
  geom_errorbar(aes(ymin=y_195, ymax=y_u95), width=0, size = 0.5, alpha = 0.2)+ # line of slope 1 through origin
  geom_abline(slope = 1, intercept = 0, color = 'black')+
  # make square
  coord_fixed(xlim = c(-4, 4), ylim = c(-4, 4))+
  theme_bw()+ 
  scale_color_viridis()+
  labs(color = "COVID Cases Per Capita (z-score)",
  y = "Predicted swing (z-score)",
  x = "Observed swing (z-score)")
```
```{r}
# save png
ggsave('figs/countyPostPred.png')
```

